Looking for a Signal in the Noise: Revisiting Obesity and the microbiome

### 2016-03-08 Sze_ObesityMicrobiomeMeta_PeerJ_2016.Rmd
Revise Methods section

*******
##### Abstract
**Rationale:** The hypothesis that changes in the microbiome are associated with obesity has been widely accepted. However, two recent studies re-analyzed published data and found that when the datasets were analyzed independently they could not support the hypothesis. In this meta analysis, we retested the hypothesis by increasing the number of data sets examined and investigating the effect of pooling the results across the individual datasets.

@@

*******
##### Introduction
Obesity is a growing worldwide health concern with approximately 20% of the youth (aged 2-19) in the United States classified as either overweight or obese [@ogden_prevalence_2014]. This number increases to approximately 35% in adults (aged 20 or older) and these statistics have seen little change since 2003 [@ogden_prevalence_2014]. Traditionally the body mass index (BMI) has been used as the traditional method of classifying individuals as non-obese or obese [@lichtash_body_2013]. Recently, there has been increased interest in the role of the microbiome in modulating obesity [@brahe_can_2016; @dror_microbiota_2016]. If the microbiome does affect obesity status, then manipulating the microbiome could have a significant role in the future treatment of obesity and in helping to stem the current epidemic.

@@

*******
##### **Methods:**  
*Literature Review and Study Inclusion*

We attempted to stick to the original PRISMA guidelines for performing a meta-analysis as published by Moher et al [-@moher_preferred_2010]. A detailed description of the literature seach and the exact formula we used for the initial PubMed search can be found in the online supplement. In brief, a total of 196 studies made it through the initial identification of studies. Screening removed a total of 184 studies leaving a total of 12 eligible studies for this meta-analysis. Upon a full reading of these specific manuscripts 2 were exculded and the reasons can be found in the online supplement. A total of 10 studies were ultimately deemed eligible for qualitative synthesis [Figure 1]. One final study was removed from the quantitative analysis since no individual in the study was found to be able to be classified in the obese category based on the classical cutoffs. A total of 3 studies were from records identified by database searching [@turnbaugh_core_2009; @zupancic_analysis_2012; @escobar_gut_2014] and the other 6 were from records identified by other sources [@ross_16s_2015; @arumugam_enterotypes_2011; @goodrich_human_2014; @wu_linking_2011; @conlan_species-level_2012]. One of these studies (Baxter) was used on a data set that was not originally designed to investigate obesity in order to help mitigate any possible publication bias associated with the bacterial microbiome and obesity.

Although there have been numerous RT-PCR and qPCR studies directed at examining the microbiome and obesity [@guo_development_2008; @schwiertz_microbiota_2010]. The main method of choice to obtain the most complete picture of the overall microbial community is through 16S sequencing of a specific hypervariable region on the 16S rRNA gene. The sequencers used to complete this task have varied depending on research group with the most popular being the Roche 454 and Illumina MiSeq. A priori we decided that for this report we would limit ourselves to analyzing studies were 16S sequencing data is avaliable along with BMI information. We also analyzed every study with a uniform approach utilizing the mothur software package for sequence processing and clean up and R statistical software for analysis.

*Subject Demographics and Study Outcomes*

The overall demographics for individuals used in this meta-analysis split by their respective study can be found in table 1. A summary of the study type, study population, whether the data set is published, and the outcomes reported can be found in table 2. We found a large variation in reported outcomes with two studies finding bacterial community associations with obesity and three studies finding some association between either Firmicutes, Bacteroidetes, or the Bacteroidetes/Firmicutes ratio and obesity. However, a total of three manuscripts found no correlations between the microbiome and obesity. This is in addition to the two re-analysis studies published in 2014. The cutoffs we used in this study for the BMI groups were lean (BMI less than or equal to 24), overweight (BMI was greater than 24 but less than 30), and obese (BMI greater than or equal to 30). For analysis involving obese and non-obese groups the cutoffs we used were a BMI less than 30 (lean/non-obese) and a BMI of equal to or greater than 30 (obese).

*Sequence Analysis Pipeline*

All data was publically available and the sequence data was downloaded from the NCBI Sequence Read Archive, MG-RAST, or the European Nucleotide Archive. A total of 6/9 [@turnbaugh_core_2009; @ross_16s_2015; @zupancic_analysis_2012; @wu_linking_2011; @escobar_gut_2014; @conlan_species-level_2012] studies were processed by staying as close as possible to the mothur 454 standard operating procedure, which can be found at http://www.mothur.org/wiki/454_SOP. A total of 2/9 [@goodrich_human_2014] of the studies were processed by staying as close as possible to the mothur MiSeq standard operating procedure found at http://www.mothur.org/wiki/MiSeq_SOP. The final data set [@arumugam_enterotypes_2011] that was used took advantage of the Metaphlan2 package [@truong_metaphlan2_2015] to generate the necessary data for this analysis. A comparison of the similarity of Metaphlan2 to 16S sequencing using the Human Microbiome Project (HMP) data set can be found in the online supplement [Figure S1-S3]. For the studies that utilized 16S rRNA gene sequencing three studies used primers targeting the V1-V3 region [@ross_16s_2015; @escobar_gut_2014; @zupancic_analysis_2012], two studies used the V4 region [@goodrich_human_2014], and one study used the V1-V2 region [@wu_linking_2011]. Where there were two choices of data sets to use we used the V3-V5 data of the HMP [@conlan_species-level_2012] and the V2 data from the Turnbaugh study [-@turnbaugh_core_2009]. A detailed walkthrough of the entire analysis pipeline can be found at https://github.com/SchlossLab/Sze_ObesityMicrobiomeMeta_PeerJ_2016/.

*Data Analysis*

We performed the analysis using R version 3.2.2 and R studio version 0.99.491. We split the overall analysis into three general categories. The first category involved following the same method employed by Finucane et al. and Walters et al. [-@finucane_taxonomic_2014; -@walters_meta-analyses_2014] in which each study was re-analyzed seperately for associations with BMI. The main associations that we explored for BMI in each seperate data set were the Bacteroidetes phyla, Firmicutes phyla, Bacteroidetes/Firmicutes (B/F) ratio, Shannon diversity, OTU richness, NMDS visualization of Bray-Curtis distance matrix PERMANOVA, and relative risk (RR). A two-way T-test was performed for comparison for non-obese versus obese. Finally, Random forest analysis to identify the best variables for classification of obese and non-obese was performed to look if there were similarities between data sets that we may have missed. The R package AUCRF (v1.1) was used for this.

For the second category of analysis we pooled all the data from each seperate study together to asses whether the combined direction of all the studies increased an individual's RR for being obese. We investigated both Shannon diversity and the B/F ratio in detail for this part. In order to generate the RR we used the median value of each study and the number of those that were non-obese or obese in each group (above or below the median respectively) was recorded and a Fisher exact-test was performed. The relative risk from the the Fisher exact-test was obtained using the epiR (v0.9-6.9) package in R. We obtained the pooled RR and confidence intervals (CI) from the metafor (v1.9-8) package in R. Finally, we also assessed the potential bias in the two pooled analysis by using a funnel plot.

The third and final category involved using  Zscore normalizations to pool all the data together to investigate the correlation between either BMI and Shannon diversity or BMI and the B/F ratio. We used a two-way t-test on the Zscore values to investigate if there was a difference between non-obese and obese individuals. In order to asses differences between the BMI categories (if data was available) we used ANOVA with a tukey post hoc test. Where appropriate we applied a Bonferroni correction for multiple comparisons. In all other cases a P-value under 0.05 was considered to be significant.

*******
##### Results:

*Seperate Analysis:*

We used the categorical variable of obese or non-obese to analyze each data set seperately for significant correlations to Bacteroidetes, Firmicutes, Bacteroidetes/Firmicute (B/F) ratio, Shannon diversity, OTU richness, evenness, and the Bray-Curtis distance matrix which was a similar approach taken by two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014]. Using a P-value cutoff for significance as less than 0.05 we found a total of 1 significant result for Shannon diversity, 2 significant results for OTU richness, and 1 signficant result for the Bray-Curtis distance matrix. There were no significant results for Bacteroidetes, Firmicutes, or the B/F ratio. The full summary of the exact P-values for each respective study and variable tested can be found in table 3.

@@

*Classical Pooled Meta-analysis:*

@@

The final forest plot that we used for this analysis investigated how the different OTUs, classified to Ruminococcaceae, that were important for the classification of obese and non-obese based on the OOB AUC in the the seperate study analysis looked when pooled together to assess the RR [Figure S4]. There was no significant increase in RR based on the Ruminococcaceae OTU (RR = 0.96 (CI 0.60, 1.52), P-value = 0.8575). Thus it would seem that although Ruminococcaceae is important for classification the directionality is study dependent [Figure S4].

*ZScore Normalization Pooled Analysis:*

Using the third approach with Zscore normalization we found that there was a significant difference in the Shannon diversity between the obese and non-obese group (P-value = 0.00493) [Figure 5A]. Further there was also a significant difference in this normalized Shannon diversity value when the data was split between the three BMI groups (P-value = 0.03826) [Figure 5B]. Using a Tukey post hoc test this difference was driven by the overweight versus the obese group (P-value = 0.03945). However, in contrast to this result we found that the B/F ratio, using the ZScore normalization, was not significant different between obese and non-obese indivdiuals (P-value = 0.4516) [Figure 5C]. Additionally, we did not observe a difference when the indivdiuals were seperated by BMI groups (P-value = 0.3317) [Figure 5D].

@@

Our meta-analysis helps to provide a little more clarity to the on going debate of whether or not there are specific microbiome variables that are influenced by obesity. Two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014] have stated that the data do not support that there is a microbiome difference between non-obese and obese individuals. However, neither of these studies made an attempt to pool the existing data together to try and harness the additional power that this would give. Here we perform an extensive literature review of the existing studies on the microbiome and obesity and perform a meta-analysis on the studies that remained based on our inclusion and exclusion criteria. We analyze the data using three different approaches. The first approach uses the method taken by the previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014], the second approach takes a meta-analysis approach using RR representation by forest plots and funnel plots, the final approach also takes a meta-analysis approach but uses ZScore normalizations and does not use RR.

@@

We do acknowledge that this study does have a few limitations that are worth mentioning. First,  it has been documented that in a number of groups BMI is not a good classification of obesity [@who_expert_consultation_appropriate_2004; @rahman_accuracy_2010]. Of particular note, the asian population has a seperate set of guidelines for obesity [@who_expert_consultation_appropriate_2004]. Although the asain population in the studies analyzed are very small [Table 1] it is possible that this may contribute to some of the noise in the data. However, it is important to keep in mind that for these measures, specifically the RR, there does not seem to be any study bias due to studies having more ethnic diversity [Figure 4]. Additionally, we are also limited in our BMI classifications by what the previous studies published and performed. In general, not enough studies have been done to investigate the microbiome in the context of other markers for obesity or on different ethnicities to be able to perform the analysis that we did. Thus as more specific studies are published that address ethnicity and studies on the microbiome that use better markers of obesity are completed it is entirely possible that the noise associated with these measurements will decrease.

@@

Fourth, although a significant result exists for the RR for both Shannon diversity and the B/F ratio, it is small. This is also the case for the ZScore normalization analysis of the Shannon diversity. As an example, a lower than median Shannon diversity only has an overall 18% increase of obesity in it versus the higher than median Shannon diversity group. So even though there is a detectable signal it remains to be seen if this is a biologically significant phenomenon that could have a tangible effect on obesity.

@@

*******
##### Tables:

**Table 1:Summary Demographics of Individuals used in the Meta-analysis.**     
```{r Table_1, echo=FALSE, message=FALSE, warning=FALSE}

@@

*Only BMI group information was provided and only an age range was given.

**Table 2:Summary of Important Study Characteristics**

@@

Outcomes <- c("N/A (Unplublished)", "Mentions obesity, no specific correlations made though", "Christensenellaceae higher in low BMI individuals. No other correlations found.", "Firmicutes less abundant at higher BMI", "Bacteroidetes/Firmicutes ratio adjusted BMI correlation", "No significant correlations", "Unweighted UniFrac Correlation with BMI", "No significant  correlations", "Decrease diversity, Bacteroidetes, and Actinobacteria correlated with increased BMI ")

@@

**Table 3:Summary of P-values for Measurements of Interest for each Individual Study for Obese versus Normal**

@@

###########################################################################
############ ZScore Data Preperation ###################################
###########################################################################

@@

#Create group analysis table to harness full data set
#For ZScore normalization
d1.obese <- as.character(data1$obese)
t.obese <- as.character(turn.data$obese)

@@

d1.H.Zscore <- data1$ZH
t.H.Zscore <- turn.data$ZH

#Combine everything together
comb1.obese <- c(d1.obese, t.obese)
comb1.bmi.class <- c(d1.bmi.class, t.bmi.class)
comb1.H.Zscore <- as.data.frame(c(d1.H.Zscore, t.H.Zscore))

group.data.test1 <- cbind(comb1.H.Zscore, comb1.obese, comb1.bmi.class)
colnames(group.data.test1) <- c("H.Zscore", "obese", "BMI.class")

@@

##Run Zscore normalized Bacter and Firm overall analysis
#data1 order Escobar, GLNE, CCHC, HMP, Korean, MetaHit
d1.lzbf.corr <- data1$ZlogBF
t.lzbf.corr <- turn.data$ZLogBF

@@

## Perform the statistical tests
combZHTtest <- t.test(H.Zscore ~ obese, data=group.data.test) #P-value = 0.006061
combZHANOVA <- anova(lm(H.Zscore ~ BMI.class, data=group.data.test)) #P-value = 0.0193
a2 <- aov(H.Zscore ~ BMI.class, data=group.data.test)
combZHTukey <- TukeyHSD(x=a2, 'BMI.class', conf.level=0.95)
combZHTukeyFinal <- combZHTukey$BMI.class

@@

```{r Figure_5, echo=FALSE, message=FALSE, warning=FALSE}
# Section A
p1 <- ggplot(group.data.test, aes(obese, H.Zscore))
p2 <- p1 + geom_boxplot(aes(fill = obese)) + scale_fill_brewer(palette="Set1")
p3 <- p2 + theme(axis.text=element_text(size=8), axis.title=element_text(size=10, face="bold"))
p4 <- p3 + xlab("Obese Category") + ylab("ZScore Normalized H")
p5 <- p4 + guides(fill=guide_legend(title="Obese")) + theme(legend.title = element_text(size=8, face="bold"))
p6 <- p5 + theme(axis.line=element_line(colour="black"),
                 panel.grid.major = element_blank(),
                 panel.grid.minor = element_blank(),
                 panel.border = element_blank(),
                 panel.background = element_blank())
p7 <- p6 + ggtitle('A') + theme(plot.title=element_text(hjust=0, face="bold", size=20))

# Section B
q1 <- ggplot(group.data.test, aes(BMI.class, H.Zscore))
q2 <- q1 + geom_boxplot(aes(fill = BMI.class)) + scale_fill_brewer(palette="Set1")
q3 <- q2 + theme(axis.text=element_text(size=8), axis.title=element_text(size=10, face="bold"))
q4 <- q3 + xlab("BMI Classification") + ylab("ZScore Normalized H")
q5 <- q4 + guides(fill=guide_legend(title="BMI Class")) + theme(legend.title = element_text(size=8, face="bold"))
q6 <- q5 + theme(axis.line=element_line(colour="black"),
                 panel.grid.major = element_blank(),
                 panel.grid.minor = element_blank(),
                 panel.border = element_blank(),
                 panel.background = element_blank())
q7 <- q6 + ggtitle('B') + theme(plot.title=element_text(hjust=0, face="bold", size=20))

# Section C
bf1 <- ggplot(group.data.test, aes(obese, ZLogBF))
bf2 <- bf1 + geom_boxplot(aes(fill = obese)) + scale_fill_brewer(palette="Set1")
bf3 <- bf2 + theme(axis.text=element_text(size=8), axis.title=element_text(size=10, face="bold"))
bf4 <- bf3 + xlab("Obese Category") + ylab("ZScore Normalized Log B/F Ratio")
bf5 <- bf4 + guides(fill=guide_legend(title="Obese")) + theme(legend.title = element_text(size=8, face="bold"))
bf6 <- bf5 + theme(axis.line=element_line(colour="black"),
                   panel.grid.major = element_blank(),
                   panel.grid.minor = element_blank(),
                   panel.border = element_blank(),
                   panel.background = element_blank())
bf7 <- bf6 + ggtitle('C') + theme(plot.title=element_text(hjust=0, face="bold", size=20))

# Section D
qbf1 <- ggplot(group.data.test, aes(BMI.class, ZLogBF))
qbf2 <- qbf1 + geom_boxplot(aes(fill = BMI.class)) + scale_fill_brewer(palette="Set1")
qbf3 <- qbf2 + theme(axis.text=element_text(size=8), axis.title=element_text(size=10, face="bold"))
qbf4 <- qbf3 + xlab("BMI Classification") + ylab("ZScore Normalized Log B/F Ratio")
qbf5 <- qbf4 + guides(fill=guide_legend(title="BMI Class")) + theme(legend.title = element_text(size=8, face="bold"))
qbf6 <- qbf5 + theme(axis.line=element_line(colour="black"),
                     panel.grid.major = element_blank(),
                     panel.grid.minor = element_blank(),
                     panel.border = element_blank(),
                     panel.background = element_blank())
qbf7 <- qbf6 + ggtitle('D') + theme(plot.title=element_text(hjust=0, face="bold", size=20))

@@

**Figure 5: Pooled Analysis of Shannon Diversity and B/F ratio by either Obesity Status or BMI Groups.**  **A)** ZScore Normalized Shannon Diversity and Obesity. There was a significant difference between non-obese (n=1081) and obese groups (n=435) (P-value = `r signif(combZHTtest$p.value, 3)`). **B)** ZScore Normalized Shannon Diversity by BMI Group. Total n for lean, overweight, and obese groups are 563, 435, 518 respectively. Using an ANOVA with tukey post-hoc testing there was a significant diference between the overweight and obese group (P-value = `r signif(combZHTukeyFinal[3, 4], 3)`) but no difference between lean and obese group (P-value = `r signif(combZHTukeyFinal[2, 4], 3)`). **C)** ZScore Normalized B/F Ratio and Obesity. There was no significant difference between the non-obese (n=1081) and obese (n=435) group based on B/F ratio (P-value = `r signif(combZLogBFTtest$p.value, 3)`). **D)** ZScore Normalized B/F Ratio by BMI Group. Total n for lean, overweight, and obese groups are 563, 435, 518 respectively. There was no significant difference between any of the three groups based on ANOVA with Tukey post-hoc testing (P-value = `r signif(combZLogBFANOVA[1, 5], 3)`).

>>>>

*******
## Abstract
**Rationale:** The hypothesis that changes in the microbiome are associated with obesity has been widely accepted. However, two recent studies re-analyzed published data and found that when the datasets were analyzed independently they could not support the hypothesis. In this meta analysis, we retested the hypothesis by increasing the number of data sets examined and investigating the effect of pooling the results across the individual datasets.

@@

\newpage

## Introduction
Obesity is a growing worldwide health concern with approximately 20% of the youth (aged 2-19) in the United States classified as either overweight or obese [@ogden_prevalence_2014]. This number increases to approximately 35% in adults (aged 20 or older) and these statistics have seen little change since 2003 [@ogden_prevalence_2014]. Traditionally the body mass index (BMI) has been used as the traditional method of classifying individuals as non-obese or obese [@lichtash_body_2013]. Recently, there has been increased interest in the role of the microbiome in modulating obesity [@brahe_can_2016; @dror_microbiota_2016]. If the microbiome does affect obesity status, then manipulating the microbiome could have a significant role in the future treatment of obesity and in helping to stem the current epidemic.

@@

## Methods
***Literature Review and Study Inclusion***
We followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines to identify studies to include in our meta-analysis [@moher_preferred_2010]. A detailed description of our selection process and the exact search terms are provided in the Supplement [Figure 1]. Briefly, we searched PubMed for original research studies that involved studying obesity and the human microbiome. The initial search yielded 187 studies. We identified nine additional studies that were not designed to explicitly test for an association between the microbiome and obesity. We then manually curated the 196 studies to select those studies that included BMI and sequence data. This yielded 10 eligible studies. An additional study was removed from our analysis because no individuals in the study had a BMI over 30. Among the final 9 studies, 3 were from identified from our PubMed search [@turnbaugh_core_2009; @zupancic_analysis_2012; @escobar_gut_2014], 5 were originally identified from the 9 studies that did not explicitly investigate obesity but included BMI data [@ross_16s_2015; @arumugam_enterotypes_2011; @goodrich_human_2014; @wu_linking_2011; @Huttenhower2012], and one dataset was used (Baxter) because at the time of our analysis it was unpublished and enabled us to mitigate against publication biases associated with the bacterial microbiome and obesity. The nine studies are summarized in Tables 1 and 2.

***Sequence Analysis Pipeline***
All sequence data were publicly available and were downloaded from the NCBI Sequence Read Archive, the European Nucleotide Archive, or the investigators' personal website. The nine studies used 454 [@turnbaugh_core_2009; @zupancic_analysis_2012; @ross_16s_2015; @escobar_gut_2014; @wu_linking_2011; @Huttenhower2012] or Illumina sequencing [@goodrich_human_2014; @arumugam_enterotypes_2011; Baxter]. All of these studies used amplification-based 16S rRNA gene sequencing except for one, which used metagenomic shotgun sequencing [@arumugam_enterotypes_2011]. Among the studies that sequenced the 16S rRNA gene, the researchers targeted the V1-V2  [@wu_linking_2011], V1-V3 [@ross_16s_2015; @escobar_gut_2014; @zupancic_analysis_2012], V3-V5 [@Huttenhower2012], and V4 [@goodrich_human_2014] regions. For those studies where multiple regions were sequenced,  we selected the region that corresponded to the largest number of subjects [@turnbaugh_core_2009; @Huttenhower2012]. We processed the 16S rRNA gene sequence data as described in the original studies. The taxonomic data from the metagenomic shotgun sequence data were obtained using Metaphlan2 [@truong_metaphlan2_2015]. A detailed and reproducible description of how the data were processed can be found at https://github.com/SchlossLab/Sze_ObesityMicrobiomeMeta_PeerJ_2016/.

***Data Analysis***
We performed the analysis using R (`r paste0(R.Version()$major,'.', R.Version()$minor)`). We split the overall analysis into three general strategies. In the first strategy we followed the approach employed by Finucane et al [-@finucane_taxonomic_2014] and Walters et al [-@walters_meta-analyses_2014] where each study was re-analyzed separately to identify associations between BMI and the relative abundance of Bacteroidetes and Firmicutes, the ratio of Bacteroidetes and Firmicutes relative abundances (B/F), Shannon diversity, and number of OTUs. A two-tailed Wilcoxon test was performed for comparison for non-obese and obese individuals. We also measured the Spearman correlation coefficient between each metric and the actual BMI value. Next, we compared the community structure from non-obese and obese individuals using PERMANOVA analysis of Bray-Curtis distance matrices. This analysis was performed using the vegan R package. Finally, we used the AUCRF R package to implement a Random Forest analysis to identify the best OTUs and alpha-diversity metrics for distinguish non-obese and obese individuals.

In the second strategy, we pooled the relative risk (RR) values for each study. For each study we partitioned the subjects into a low or high group depending on whether their diversity or B/F was below or above the median value for the study. The RR was then calculated as the ratio of the number of obese individuals in the low group to the number of obese individuals in the high group. We then performed a Fisher exact-test to test whether the RR was significantly different from 1.0 using the epiR (v0.9-6.9) and metafor (v1.9-8) packages. Finally, we assessed the potential publication bias in the diversity and B/F analyses using a funnel plot.

In the third strategy, we normalized the diversity and B/F data within each study and pooled the normalized data across studies. We applied a Z-score normalization to the Shannon diversity data and log-transformed B/F data within each study to have a mean of zero and standard deviation of one. We then pooled Z-scores across studies and used a two-tailed T-test on the values to test for a significant difference in Z-scores among non-obese and obese individuals. To assess differences between the lean, overweight, and obese BMI categories we used ANOVA with a Tukey post hoc test. We also measured the Spearman correlation coefficient between the Z-scores for each metric and BMI values.

## Results:

***Separate Analysis***
We used the categorical variable of obese or non-obese to analyze each data set separately for significant correlations to Bacteroidetes, Firmicutes, Bacteroidetes/Firmicute (B/F) ratio, Shannon diversity, OTU richness, evenness, and the Bray-Curtis distance matrix which was a similar approach taken by two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014]. Using a P-value cutoff for significance as less than 0.05 we found a total of 1 significant result for Shannon diversity, 2 significant results for OTU richness, and 1 signficant result for the Bray-Curtis distance matrix. There were no significant results for Bacteroidetes, Firmicutes, or the B/F ratio. The full summary of the exact P-values for each respective study and variable tested can be found in table 3.

@@

***Classical Pooled Meta-analysis***

@@

The final forest plot that we used for this analysis investigated how the different OTUs, classified to Ruminococcaceae, that were important for the classification of obese and non-obese based on the OOB AUC in the the separate study analysis looked when pooled together to assess the RR [Figure S4]. There was no significant increase in RR based on the Ruminococcaceae OTU (RR = 0.96 (CI 0.60, 1.52), P-value = 0.8575). Thus it would seem that although Ruminococcaceae is important for classification the directionality is study dependent [Figure S4].

***Z-Score Normalization Pooled Analysis***

Using the third approach with Z-score normalization we found that there was a significant difference in the Shannon diversity between the obese and non-obese group (P-value = 0.00493) [Figure 5A]. Further there was also a significant difference in this normalized Shannon diversity value when the data was split between the three BMI groups (P-value = 0.03826) [Figure 5B]. Using a Tukey post hoc test this difference was driven by the overweight versus the obese group (P-value = 0.03945). However, in contrast to this result we found that the B/F ratio, using the Z-score normalization, was not significant different between obese and non-obese individuals (P-value = 0.4516) [Figure 5C]. Additionally, we did not observe a difference when the individuals were separated by BMI groups (P-value = 0.3317) [Figure 5D].

@@

Our meta-analysis helps to provide a little more clarity to the on going debate of whether or not there are specific microbiome variables that are influenced by obesity. Two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014] have stated that the data do not support that there is a microbiome difference between non-obese and obese individuals. However, neither of these studies made an attempt to pool the existing data together to try and harness the additional power that this would give. Here we perform an extensive literature review of the existing studies on the microbiome and obesity and perform a meta-analysis on the studies that remained based on our inclusion and exclusion criteria. We analyze the data using three different approaches. The first approach uses the method taken by the previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014], the second approach takes a meta-analysis approach using RR representation by forest plots and funnel plots, the final approach also takes a meta-analysis approach but uses Z-score normalizations and does not use RR.

@@

We do acknowledge that this study does have a few limitations that are worth mentioning. First,  it has been documented that in a number of groups BMI is not a good classification of obesity [@who_expert_consultation_appropriate_2004; @rahman_accuracy_2010]. Of particular note, the asian population has a separate set of guidelines for obesity [@who_expert_consultation_appropriate_2004]. Although the asain population in the studies analyzed are very small [Table 1] it is possible that this may contribute to some of the noise in the data. However, it is important to keep in mind that for these measures, specifically the RR, there does not seem to be any study bias due to studies having more ethnic diversity [Figure 4]. Additionally, we are also limited in our BMI classifications by what the previous studies published and performed. In general, not enough studies have been done to investigate the microbiome in the context of other markers for obesity or on different ethnicities to be able to perform the analysis that we did. Thus as more specific studies are published that address ethnicity and studies on the microbiome that use better markers of obesity are completed it is entirely possible that the noise associated with these measurements will decrease.

@@

Fourth, although a significant result exists for the RR for both Shannon diversity and the B/F ratio, it is small. This is also the case for the Z-score normalization analysis of the Shannon diversity. As an example, a lower than median Shannon diversity only has an overall 18% increase of obesity in it versus the higher than median Shannon diversity group. So even though there is a detectable signal it remains to be seen if this is a biologically significant phenomenon that could have a tangible effect on obesity.

@@

\newpage

## Tables

**Table 1. Summary Demographics of Individuals used in the Meta-analysis.**
```{r Table_1, echo=FALSE, message=FALSE, warning=FALSE}

@@

* Only BMI group information was provided and only an age range was given.

\newpage

**Table 2. Summary of Important Study Characteristics.**

@@

Outcomes <- c("N/A (Unpublished)", "Mentions obesity, no specific correlations made though", "Christensenellaceae higher in low BMI individuals. No other correlations found.", "Firmicutes less abundant at higher BMI", "Bacteroidetes/Firmicutes ratio adjusted BMI correlation", "No significant correlations", "Unweighted UniFrac Correlation with BMI", "No significant  correlations", "Decrease diversity, Bacteroidetes, and Actinobacteria correlated with increased BMI ")

@@

**Table 3. Summary of P-values for Measurements of Interest for each Individual Study for Obese versus Normal**

@@

###########################################################################
############ Z-score Data Preparation ###################################
###########################################################################

@@

#Create group analysis table to harness full data set
#For Z-score normalization
d1.obese <- as.character(data1$obese)
t.obese <- as.character(turn.data$obese)

@@

d1.H.Z-score <- data1$ZH
t.H.Z-score <- turn.data$ZH

#Combine everything together
comb1.obese <- c(d1.obese, t.obese)
comb1.bmi.class <- c(d1.bmi.class, t.bmi.class)
comb1.H.Z-score <- as.data.frame(c(d1.H.Z-score, t.H.Z-score))

group.data.test1 <- cbind(comb1.H.Z-score, comb1.obese, comb1.bmi.class)
colnames(group.data.test1) <- c("H.Z-score", "obese", "BMI.class")

@@

##Run Z-score normalized Bacter and Firm overall analysis
#data1 order Escobar, GLNE, CCHC, HMP, Korean, MetaHit
d1.lzbf.corr <- data1$ZlogBF
t.lzbf.corr <- turn.data$ZLogBF

@@

## Perform the statistical tests
combZHTtest <- t.test(H.Z-score ~ obese, data=group.data.test) #P-value = 0.006061
combZHANOVA <- anova(lm(H.Z-score ~ BMI.class, data=group.data.test)) #P-value = 0.0193
a2 <- aov(H.Z-score ~ BMI.class, data=group.data.test)
combZHTukey <- TukeyHSD(x=a2, 'BMI.class', conf.level=0.95)
combZHTukeyFinal <- combZHTukey$BMI.class

@@

```{r Figure_5, echo=FALSE, message=FALSE, warning=FALSE}
# Section A
p1 <- ggplot(group.data.test, aes(obese, H.Z-score))
p2 <- p1 + geom_boxplot(aes(fill = obese)) + scale_fill_brewer(palette="Set1")
p3 <- p2 + theme(axis.text=element_text(size=8), axis.title=element_text(size=10, face="bold"))
p4 <- p3 + xlab("Obese Category") + ylab("Z-score Normalized H")
p5 <- p4 + guides(fill=guide_legend(title="Obese")) + theme(legend.title = element_text(size=8, face="bold"))
p6 <- p5 + theme(axis.line=element_line(colour="black"),
                 panel.grid.major = element_blank(),
                 panel.grid.minor = element_blank(),
                 panel.border = element_blank(),
                 panel.background = element_blank())
p7 <- p6 + ggtitle('A') + theme(plot.title=element_text(hjust=0, face="bold", size=20))

# Section B
q1 <- ggplot(group.data.test, aes(BMI.class, H.Z-score))
q2 <- q1 + geom_boxplot(aes(fill = BMI.class)) + scale_fill_brewer(palette="Set1")
q3 <- q2 + theme(axis.text=element_text(size=8), axis.title=element_text(size=10, face="bold"))
q4 <- q3 + xlab("BMI Classification") + ylab("Z-score Normalized H")
q5 <- q4 + guides(fill=guide_legend(title="BMI Class")) + theme(legend.title = element_text(size=8, face="bold"))
q6 <- q5 + theme(axis.line=element_line(colour="black"),
                 panel.grid.major = element_blank(),
                 panel.grid.minor = element_blank(),
                 panel.border = element_blank(),
                 panel.background = element_blank())
q7 <- q6 + ggtitle('B') + theme(plot.title=element_text(hjust=0, face="bold", size=20))

# Section C
bf1 <- ggplot(group.data.test, aes(obese, ZLogBF))
bf2 <- bf1 + geom_boxplot(aes(fill = obese)) + scale_fill_brewer(palette="Set1")
bf3 <- bf2 + theme(axis.text=element_text(size=8), axis.title=element_text(size=10, face="bold"))
bf4 <- bf3 + xlab("Obese Category") + ylab("Z-score Normalized Log B/F Ratio")
bf5 <- bf4 + guides(fill=guide_legend(title="Obese")) + theme(legend.title = element_text(size=8, face="bold"))
bf6 <- bf5 + theme(axis.line=element_line(colour="black"),
                   panel.grid.major = element_blank(),
                   panel.grid.minor = element_blank(),
                   panel.border = element_blank(),
                   panel.background = element_blank())
bf7 <- bf6 + ggtitle('C') + theme(plot.title=element_text(hjust=0, face="bold", size=20))

# Section D
qbf1 <- ggplot(group.data.test, aes(BMI.class, ZLogBF))
qbf2 <- qbf1 + geom_boxplot(aes(fill = BMI.class)) + scale_fill_brewer(palette="Set1")
qbf3 <- qbf2 + theme(axis.text=element_text(size=8), axis.title=element_text(size=10, face="bold"))
qbf4 <- qbf3 + xlab("BMI Classification") + ylab("Z-score Normalized Log B/F Ratio")
qbf5 <- qbf4 + guides(fill=guide_legend(title="BMI Class")) + theme(legend.title = element_text(size=8, face="bold"))
qbf6 <- qbf5 + theme(axis.line=element_line(colour="black"),
                     panel.grid.major = element_blank(),
                     panel.grid.minor = element_blank(),
                     panel.border = element_blank(),
                     panel.background = element_blank())
qbf7 <- qbf6 + ggtitle('D') + theme(plot.title=element_text(hjust=0, face="bold", size=20))

@@

**Figure 5: Pooled Analysis of Shannon Diversity and B/F ratio by either Obesity Status or BMI Groups.**  **A)** Z-score Normalized Shannon Diversity and Obesity. There was a significant difference between non-obese (n=1081) and obese groups (n=435) (P-value = `r signif(combZHTtest$p.value, 3)`). **B)** Z-score Normalized Shannon Diversity by BMI Group. Total n for lean, overweight, and obese groups are 563, 435, 518 respectively. Using an ANOVA with tukey post-hoc testing there was a significant diference between the overweight and obese group (P-value = `r signif(combZHTukeyFinal[3, 4], 3)`) but no difference between lean and obese group (P-value = `r signif(combZHTukeyFinal[2, 4], 3)`). **C)** Z-score Normalized B/F Ratio and Obesity. There was no significant difference between the non-obese (n=1081) and obese (n=435) group based on B/F ratio (P-value = `r signif(combZLogBFTtest$p.value, 3)`). **D)** Z-score Normalized B/F Ratio by BMI Group. Total n for lean, overweight, and obese groups are 563, 435, 518 respectively. There was no significant difference between any of the three groups based on ANOVA with Tukey post-hoc testing (P-value = `r signif(combZLogBFANOVA[1, 5], 3)`).
